---
title: 'ESM 244 Lab 7 Part 2: Spatial kriging'
author: "Sophia Leiker"
date: "2/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
library(gstat)
library(stars)
```

## Read in the California depth to groundwater data

**Data source:** California Department of Water Resources. (2016). Groundwater Depth, California, Fall 2013. California Department of Water Resources. Available at: http://purl.stanford.edu/py212tb3061.

These data are measurements for depth to groundwater recorded in 2013. We'll read in the layer with `read_sf()`, check the CRS, then make a quick plot to start

```{r}

# Read in the CA county data (TIGER shapefile):
ca_counties_sf <- read_sf(here("data/ca_counties"), layer = "CA_Counties_TIGER2016") %>% 
  janitor::clean_names() %>% 
  select(name)

# Check the projection
st_crs(ca_counties_sf) #WGS 84

# Read in the depth to groundwater layer: 
ca_dgw_sf <- read_sf(here("data/ca_dgw/F2013_DBGS_Points_20150720_093252.gpkg")) %>% 
  janitor::clean_names()

# Check the projection:
st_crs(ca_dgw_sf) # WGS 84, epsg:4326

ca_dgw_sf <- st_transform(ca_dgw_sf, st_crs(ca_counties_sf))

# Make a quick ggplot:
ggplot() +
  geom_sf(data = ca_counties_sf) +
  geom_sf(data = ca_dgw_sf, aes(color = dgbs))
```

## Map these interactively to explore further

```{r}
tmap_mode("view")

tm_shape(ca_dgw_sf) +
  tm_dots("dgbs", palette = 'Blues')
### note - no CA map here, Leaflet does that for us

### check out all the palettes:
# tmaptools::palette_explorer() ### need package shinyjs
```

Let's just look at depth to groundwater (ft) in and around San Joaquin County. 

```{r}
# Just get sj_county
sj_county_sf <- ca_counties_sf %>% 
  filter(name == "San Joaquin")

# Now keep observations for groundwater depth within that county:
sj_depth_sf <- ca_dgw_sf %>% 
  st_intersection(sj_county_sf) #we should do biodiversity by country, so you use this intersection tool as what you feed in for a drop down menu and then it will spit out a map with the public lands vector/biodiveristy raster/land cover raster/output raster 

# Make a quick plot of those: 
ggplot() +
  geom_sf(data = sj_county_sf) +
  geom_sf(data = sj_depth_sf, aes(color = dgbs)) +
  theme_void() +
  scale_color_viridis_c()
```

### Check for duplicate locations:

Interpolation chokes when observations at identical location - spatial singularity!

```{r}
well_duplicates <- sj_depth_sf %>% 
  janitor::get_dupes(latitude, longitude) #this gets the duplicates from latitude and longitude columns (if they are both simultanously duplicated)

well_duplicates

# This will cause problems (0 spatial variance). We're just going to get rid of these problematic wells recorded at the same location: 

sj_depth_sf <- sj_depth_sf %>% 
  filter(!local_well %in% well_duplicates$local_well)

# Check again for duplicates: 
sj_depth_sf %>% 
  janitor::get_dupes(latitude, longitude) 

# No more duplicates
```

## Make a variogram: 

```{r}
# Create the variogram
sj_dgw_vgm <- gstat::variogram(dgbs ~ 1, data = sj_depth_sf)
#dgbs as a function of 1, this is calculating an intercept as the mean based on the data, this is like ordinary kriging, this is a constant mean but we do not know what the constant is

# Plot it
plot(sj_dgw_vgm) # Looks variogramish! Increasing variance as observations get further apart. Let's model this...

# Now create the variogram model, & see ?vgm for other model options: 
sj_dgw_vgm_fit <- fit.variogram(sj_dgw_vgm, 
                                model = vgm(model = "Gau")) ### default starting (there are other models such as sperical, how would one select a certain model to use and does it influence outputs significantly?)
sj_dgw_vgm_fit <- fit.variogram(sj_dgw_vgm, 
                                model = vgm(model = "Gau", ### (leave with defaults first)
                                            nugget = 20, sill = 3000, range = 30000))

# Plot them together:
plot(sj_dgw_vgm, sj_dgw_vgm_fit) 

# You can check the sum of squares of residuals for each: 
attr(sj_dgw_vgm_fit, 'SSErr') # ...and you could compare to other models...

```

## Spatial kriging

```{r}

# Make a grid over which we'll krige: 
sj_grid <- st_bbox(sj_county_sf) %>% #this is creating a bounding box to capture the box
  st_as_stars(dx = 1000, dy = 1000) %>% #using this to set up a grid, 1000m X 1000x (like a fishnet in Arc) 
  st_set_crs(st_crs(sj_county_sf)) %>% #this is setting the crs to match the sj_county_sf
  st_crop(sj_county_sf) #this is cropping the grid so it only includes points within the outline of this polygon

sj_grid

# Look at the grid we made: 
plot(sj_grid)

# Perform ordinary kriging: 
sj_dgw_krige <- krige(dgbs ~ 1, sj_depth_sf, sj_grid, model = sj_dgw_vgm_fit)
#this sj_depth_sf is the points, sj_grid is the grid we are using, the model we are using the variogram model fit that we created above

# Initial plot of kriging outcome: 
plot(sj_dgw_krige)

# Convert it to a spatial data frame
krige_df <- as.data.frame(sj_dgw_krige) %>% 
  st_as_sf(coords = c("x","y")) %>% #this is a function that converts things to a simple features object
  drop_na(var1.pred) #check to make sure this works by running class(krige_df) and it should be a sf (simple features object)

st_crs(krige_df) <- 3857

# Then we can use ggplot: 
ggplot(data = krige_df) +
  geom_sf(aes(color = var1.pred)) + #color is predicted value
  scale_color_gradient(low = "blue", high = "yellow")
```

...& there are ways to convert this to raster data / aggregate if you want...

## End variogram & kriging demo

